---
title: "Whole game - Data visualization"
output: html_notebook
---

# Whole game

## 1. Data visualization

With ggplot2

```{r}
# Loading the package

library(tidyverse)

# Data sets we will be using

library(palmerpenguins)
library(ggthemes)
```

### Penguins data frame

**Our ultimate job is to display the relationship between flipper lengths and body masses, taking into consideration the species of the penguin**

Variable -> Quantity, quality or property you can measure (variable refers to an attribute that all penguins have) 

 Value -> state of a variable

Oberservation ->  set of measurements made under similar conditions (all atributes of a single penguin) 

Tabular data -> set of values, each associated with a variable and an observation


```{r}
penguins
```

Variables as columns

```{r}
glimpse(penguins)
```
### Creating a ggplot

- mapping -> how variables are going to be mapped

- aes() -> define x and y axes

- geom -> geometrical object used to represent data

- color (inside of mapping) -> color based on the mapped points

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) + 
  geom_point(mapping = aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape =  "Species"
  ) +
  
  scale_color_colorblind()

```

*As the flipper length increases, so does the body mass. It is failry linear and strong*

### Exercises

**How many rows are in penguins? How many columns?**

344 rows and 8 columns

**What does the bill_depth_mm variable in the penguins data frame describe? Read the help for ?penguins to find out.**

It describes as "a number denoting bill depth (millimeters)"

**Make a scatterplot of bill_depth_mm vs. bill_length_mm. That is, make a scatterplot with bill_depth_mm on the y-axis and bill_length_mm on the x-axis. Describe the relationship between these two variables.**

```{r}
ggplot(
  data = penguins,
  aes(x = bill_length_mm, y = bill_depth_mm)
       ) + 
  geom_point(color="salmon") +
  labs(
    title = "Bill length and bill depth"
  )
```

For which ever value bill depth takes, the bill length will stay the same

**What happens if you make a scatterplot of species vs. bill_depth_mm? What might be a better choice of geom?**

```{r}
ggplot(
  penguins,
  aes(x = bill_depth_mm, y = species)
) +
  geom_boxplot(aes(color = species)) +
  labs(
    title = "Species and bill depth",
    color = "Species"
  )
```


**Why does the following give an error and how would you fix it?**


```{r}
ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm, y = bill_depth_mm)
) + 
  geom_point()
```

**What does the na.rm argument do in geom_point()? What is the default value of the argument? Create a scatterplot where you successfully use this argument set to TRUE.**

The na.rm = TRUE removes the warnings regarding the lack of data in a concrete observation

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm, y = bill_depth_mm)
) + 
  geom_point(na.rm = TRUE)
```


**Add the following caption to the plot you made in the previous exercise: “Data come from the palmerpenguins package.” Hint: Take a look at the documentation for labs().**

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm, y = bill_depth_mm)
) + 
  geom_point(na.rm = TRUE) +
  labs(
    subtitle = "Data come from the palmerpenguins package."
  )
```


**Recreate the following visualization. What aesthetic should bill_depth_mm be mapped to? And should it be mapped at the global level or at the geom level?**

```{r}
ggplot(
  penguins,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = bill_depth_mm)) +
  geom_smooth(mthod = "lm")
```

**Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions**

It's going to be a scatterplot, coloing the points based on from which island are the penguins and there will be some kind of regression line

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = island)
) +
  geom_point() +
  geom_smooth(se = FALSE)
```

**VWill these two graphs look different? Why/why not?**

```{r}
ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  )

```

### Visualizing

Bar charts -> used for categorical variables

Fct_infreq orders the levels based on the frequency.
```{r}
ggplot(penguins, aes(x = fct_infreq(species))) +
  geom_bar()
```

Histogram -> Takes quantitative variables

Binwidth determines the width of the interval in a histogram
```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200)
```

Density plot -> Quantitative variables (a smoothed out version of a hitogram)

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_density()
```

### Exercises

**Make a bar plot of species of penguins, where you assign species to the y aesthetic. How is this plot different?**

```{r}
penguins |>
  ggplot(aes(y=species)) +
  geom_bar()
```
This plot is different because it is plotted horizontally

**How are the following two plots different? Which aesthetic, color or fill, is more useful for changing the color of bars?**

The fill one colors in the bars and the color one just the outside of the bars

**What does the bins argument in geom_histogram() do?**

It determines the width of each one of the bars.

**Make a histogram of the carat variable in the diamonds dataset that is available when you load the tidyverse package. Experiment with different binwidths. What binwidth reveals the most interesting patterns?**

```{r}
ggplot(diamonds, aes(x = carat)) + 
  geom_histogram(bindwidth = 0.01)
```

### Visualizing relationships

**Box plot** -> 
- useful for outliers
-  Interquartile range: from 25th percentile to 75th percentile
- Middle of the box is the median (50th percentile)
- Edge of the box is 1.5 times de IQR
- Whiskers fo to the farthest non-outlier point in the distribution

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot(na.rm = TRUE)
```

They are similar to density plots

```{r}
ggplot(penguins, aes(x = body_mass_g, color = species, fill=species)) +
  geom_density(alpha=0.5, na.rm=TRUE)
```

**Bar graphs** ->

- For two categorical variables

```{r}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(na.rm=TRUE)
```

```{r}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "fill")
```

**Scatterplot** ->

- Visualizes the relationship between two numerical values

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(na.rm=TRUE)
```

- Can be also used for three or more variables

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island), na.rm=TRUE)
```

It can also be divided into facets. THe argument in facets has to be categorical.

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species), na.rm=TRUE) +
  facet_wrap(~island)
```

### Exercises

**The mpg data frame that is bundled with the ggplot2 package contains 234 observations collected by the US Environmental Protection Agency on 38 car models. Which variables in mpg are categorical? Which variables are numerical? (Hint: Type ?mpg to read the documentation for the dataset.) How can you see this information when you run mpg?**

```{r}
?mpg
```

Categorical variables in mpg: manufacturer, model, year, trans, drv, fl and class. The rest are numercial.

**ake a scatterplot of hwy vs. displ using the mpg data frame. Next, map a third, numerical variable to color, then size, then both color and size, then shape. How do these aesthetics behave differently for categorical vs. numerical variables?**

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
geom_point(aes(color=fl), na.rm=TRUE)
```


```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
geom_point(aes(size=fl), na.rm=TRUE)
```

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
geom_point(aes(color = fl, size=fl), na.rm=TRUE)
```

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
geom_point(aes(shape=fl), na.rm=TRUE)
```

**In the scatterplot of hwy vs. displ, what happens if you map a third variable to linewidth?**

```{r}
ggplot(mpg,
       aes(x = displ, y = hwy)) +
geom_point(aes(linewidth= cyl), na.rm=TRUE)
```
Nothing changes because there is no line to alter

**What happens if you map the same variable to multiple aesthetics?**

```{r}
ggplot(mpg, aes(x = hwy, y = hwy, color = hwy)) + 
  geom_point()
```
It is good to visualize that variable from multiple perspectives but it is not useful.

**Make a scatterplot of bill_depth_mm vs. bill_length_mm and color the points by species. What does adding coloring by species reveal about the relationship between these two variables? What about faceting by species?**

```{r}
ggplot(
  penguins,
  aes(x = bill_length_mm, y = bill_depth_mm)) +
    geom_point(aes(color=species), na.rm=TRUE)
```

The graph shows that as the bill_length_mm increases, the bill_depth_mm increases. THe Adelie species tend to have higher bill depth while Gentoo have longer bill length.

```{r}
ggplot(
  penguins,
  aes(x = bill_length_mm, y = bill_depth_mm)) +
    geom_point(aes(color=species), na.rm=TRUE) +
  facet_wrap(~species)
```
**Why does the following yield two separate legends? How would you fix it to combine the two legends?**

```{r}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm, y = bill_depth_mm, 
    color = species, shape = species
  )
) +
  geom_point(na.rm=TRUE) +
  labs(color = "Species", shape="Species")
```

**Create the two following stacked bar plots. Which question can you answer with the first one? Which question can you answer with the second one?**

```{r}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "fill")


ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = "fill")
```

The first one answers what is the ratio of species in each one of the islands. And the second one shows from each one of the species what is the ratio of islands

### Saving your plots

```{r}
ggplot(
  penguins,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point()
ggsave(filename = "penguin-plot.png")
```
### Exercises

**Run the following lines of code. Which of the two plots is saved as mpg-plot.png? Why?**

The one saved is the last one (cty and hwy) because it always saves the last one executed

**What do you need to change in the code above to save the plot as a PDF instead of a PNG? How could you find out what types of image files would work in ggsave()?**

To find more about the image file we could execute: ?ggsave()

```{r}
?ggsave()
```

