---
title: "Whole game - Data visualization"
output: html_notebook
---

# Whole game

## 1. Data visualization

With ggplot2

```{r}
# Loading the package

library(tidyverse)

# Data sets we will be using

library(palmerpenguins)
library(ggthemes)
```

### Penguins data frame

**Our ultimate job is to display the relationship between flipper lengths and body masses, taking into consideration the species of the penguin**

Variable -> Quantity, quality or property you can measure (variable refers to an attribute that all penguins have) 

 Value -> state of a variable

Oberservation ->  set of measurements made under similar conditions (all atributes of a single penguin) 

Tabular data -> set of values, each associated with a variable and an observation


```{r}
penguins
```

Variables as columns

```{r}
glimpse(penguins)
```
### Creating a ggplot

- mapping -> how variables are going to be mapped

- aes() -> define x and y axes

- geom -> geometrical object used to represent data

- color (inside of mapping) -> color based on the mapped points

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) + 
  geom_point(mapping = aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape =  "Species"
  ) +
  
  scale_color_colorblind()

```

*As the flipper length increases, so does the body mass. It is failry linear and strong*

### Exercises

**How many rows are in penguins? How many columns?**

344 rows and 8 columns

**What does the bill_depth_mm variable in the penguins data frame describe? Read the help for ?penguins to find out.**

It describes as "a number denoting bill depth (millimeters)"

**Make a scatterplot of bill_depth_mm vs. bill_length_mm. That is, make a scatterplot with bill_depth_mm on the y-axis and bill_length_mm on the x-axis. Describe the relationship between these two variables.**

```{r}
ggplot(
  data = penguins,
  aes(x = bill_length_mm, y = bill_depth_mm)
       ) + 
  geom_point(color="salmon") +
  labs(
    title = "Bill length and bill depth"
  )
```

For which ever value bill depth takes, the bill length will stay the same

**What happens if you make a scatterplot of species vs. bill_depth_mm? What might be a better choice of geom?**

```{r}
ggplot(
  penguins,
  aes(x = bill_depth_mm, y = species)
) +
  geom_boxplot(aes(color = species)) +
  labs(
    title = "Species and bill depth",
    color = "Species"
  )
```


**Why does the following give an error and how would you fix it?**


```{r}
ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm, y = bill_depth_mm)
) + 
  geom_point()
```

**What does the na.rm argument do in geom_point()? What is the default value of the argument? Create a scatterplot where you successfully use this argument set to TRUE.**

The na.rm = TRUE removes the warnings regarding the lack of data in a concrete observation

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm, y = bill_depth_mm)
) + 
  geom_point(na.rm = TRUE)
```


**Add the following caption to the plot you made in the previous exercise: “Data come from the palmerpenguins package.” Hint: Take a look at the documentation for labs().**

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm, y = bill_depth_mm)
) + 
  geom_point(na.rm = TRUE) +
  labs(
    subtitle = "Data come from the palmerpenguins package."
  )
```


**Recreate the following visualization. What aesthetic should bill_depth_mm be mapped to? And should it be mapped at the global level or at the geom level?**

```{r}
ggplot(
  penguins,
  aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = bill_depth_mm)) +
  geom_smooth(mthod = "lm")
```

**Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions**

It's going to be a scatterplot, coloing the points based on from which island are the penguins and there will be some kind of regression line

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = island)
) +
  geom_point() +
  geom_smooth(se = FALSE)
```

**VWill these two graphs look different? Why/why not?**

```{r}
ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  )

```

### Visualizing

Bar charts -> used for categorical variables

Fct_infreq orders the levels based on the frequency.
```{r}
ggplot(penguins, aes(x = fct_infreq(species))) +
  geom_bar()
```

Histogram -> Takes quantitative variables

Binwidth determines the width of the interval in a histogram
```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200)
```

Density plot -> Quantitative variables (a smoothed out version of a hitogram)

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_density()
```

### Exercises

**Make a bar plot of species of penguins, where you assign species to the y aesthetic. How is this plot different?**

```{r}
penguins |>
  ggplot(aes(y=species)) +
  geom_bar()
```
This plot is different because it is plotted horizontally

**How are the following two plots different? Which aesthetic, color or fill, is more useful for changing the color of bars?**

The fill one colors in the bars and the color one just the outside of the bars

**What does the bins argument in geom_histogram() do?**

It determines the width of each one of the bars.

**Make a histogram of the carat variable in the diamonds dataset that is available when you load the tidyverse package. Experiment with different binwidths. What binwidth reveals the most interesting patterns?**

```{r}
ggplot(diamonds, aes(x = carat)) + 
  geom_histogram(bindwidth = 0.01)
```

