---
title: "Data transformation"
author: '@mortellss'
date: "2025-03-05"
output: html_document
---

Sometimes we want to: create new variables, rename ones that exist, reorder observations

```{r}
library(nycflights13)
library(tidyverse)
```


The nycflights13 package contains flights departed from NYC in 2013. The data comes from the Bureau of Transportation Statistics. This is a *tibble*. The difference between a tibble and a data frame is that tibbles are designed for large datasets so they only show the first few rows and only the columns that fit on one screen.

To see the data frames:
 
- View(flights) # Open a diffeerent tab with the data frame
- print(flights, width = Inf)  # To see all columns
- glimpse()



```{r}
flights
```

```{r}
?flights
```

## Rows

filter() -> changes which rows are present without changing their order
arrange() -> changes the order of rows without changing which are present
distinct() -> find rows with unique values

**filter()**

Changes the rows that are present without chaning their order

```{r}
# Flights that departed more than 120 minutes late
flights |>
  filter(dep_delay > 120)

# This is the same as:
# filter(flights, dep_delay > 120)
```

```{r}
# Flights that departed on January 1
flights |>
  filter(month == 1 & day == 1)

# Flights that departed on January or February

flights |>
  filter(month == 1 | month == 2)

# Another way is: 
# flights |>
#   filter(month %in% c(1, 2))


# You can also save it in a variable

jan1 <- flights |>
  filter(month == 1 & day == 1)


```

**arrange()**

Changes the order of rows based on the value of the columns.

```{r}
# This is in ascending order

flights |>
  arrange(year, month, day, dep_time)

# Descending order.
# ! Only allows one argument

flights |>
  arrange(desc(dep_time))
```
**distinct()**

Find all unique rows in dataet.

```{r}
flights |>
  distinct()

# Find all unique origin and destination pairs
# Just prints a table with that unique pairs
flights |> 
  distinct(origin, dest)


# .keep_all = TRUE
# 	Prints the table with those unique pairs but shows the entirety of the column

flights |>
  distinct(origin, dest, .keep_all = TRUE)
```

```{r}
# Find the number of ocurrences in that distinct
# The sort = TRUE, sorts it in descendent order

flights |>
  count(origin, dest, sort = TRUE)
```

**mutate()**

Creates a new variable

**nrow()**

Counts the number of rows in a dataset


### Exercises

**In a single pipeline for each condition, find all flights that meet the condition: **

```{r}
# Had an arrival delay of two or more hours

flights |>
  filter(arr_delay >= 120) |>
  arrange(desc(arr_delay))

# Flew to Houston (IAH or HOU).

flights |>
  filter(dest %in% c("IAH", "HOU"))

# Were operated by United, American, or Delta.

flights |>
  filter(carrier %in% c("UA", "AA", "DL"))

# Departed in summer (July, August, and September).

flights |>
  filter(month %in% c(7, 8, 9))

# Arrived more than two hours late but didn’t leave late.
# The view() opens up another tab with the dataset

flights |> 
  filter(arr_delay >= 120 & dep_delay <= 0) |> view()

# Were delayed by at least an hour, but made up over 30 minutes in flight

flights |>
  filter(dep_delay >= 60
         & (time_hour - air_time > 30))
```


**Sort flights to find the flights with the longest departure delays. Find the flights that left earliest in the morning.**

```{r}
flights |> 
  arrange(desc(dep_delay)) |>
  arrange(sched_dep_time) |>
  relocate(dep_delay, sched_dep_time)
  
```

**Sort flights to find the fastest flights. (Hint: Try including a math calculation inside of your function.)**

```{r}
flights |>
  mutate(speed = distance / (air_time / 60)) |>
  arrange(desc(speed)) |>
  relocate(speed)
```

**Was there a flight on every day of 2013?**

```{r}
flights |>
  count(year, month, day) |>
  nrow()
```
**Which flights traveled the farthest distance? Which traveled the least distance?**

The longest distance

```{r}
flights |>
  arrange(desc(distance)) |>
  relocate(distance)
```

The shortest distance

```{r}
flights |>
  arrange(distance) |>
  relocate(distance)
```

**Does it matter what order you used filter() and arrange() if you’re using both? Why/why not? Think about the results and how much work the functions would have to do.**

It does not matter the order.

## Column

**mutate()**

Adds new columns that are calculated from the existing columns

```{r}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60
  ) |>
  relocate(gain, speed)

# This is the same as using the relocate 
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .before = 1
  )

```


We can also use the .after to add after a variable. So in the next examples the new variables gain and speed go after the variable day

```{r}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .after = day
  )
```

THe argument .used only keeps those variables involved in the mutate() step. The next example will only show: gain, hours, gain_per_hou, dep_delay, arr_delay and air_time

```{r}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used"
  )
```

**When we mutuate, the changes made are not stored in the data frame**


**select()**

Only shows those columns

```{r}
flights |>
  select(year, month, day)
```

Now,those between the two variables indicated (invlusively)

```{r}
flights |> 
  select(year:day)
```

Now all of them except the ones indicates (inclusively)

```{r}
flights |> 
  select(!year:day)
```

Select all aocumns that are character

```{r}
flights |>
  select(where(is.character))
```

Or other functions you can use within select():
- starts_with("abc")
- ends_with("xyz")
- contains("ijk")
- num_range("x", 1:3)
- matches()

You can also rename variables as you select:

```{r}
flights |> 
  select(tail_num = tailnum)
```


**rename()**

The rename method in select only shows you that specific column but with the rename function it keeps all the existing variables

```{r}
flights |> 
  rename(tail_num = tailnum)
```

**relocate()**

Moves certain variables to the front

```{r}
flights |> 
  relocate(time_hour, air_time)
```

You can also combine the .after and .before with this function

```{r}
# Relocates are the variables from year to dep_time after time_hour
flights |> 
  relocate(year:dep_time, .after = time_hour)

# Relocates all those variables starting with "arr" after dep_time

flights |> 
  relocate(starts_with("arr"), .before = dep_time)
```

### Exercises

**Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?**

```{r}
flights |> 
  select(dep_time, sched_dep_time, dep_delay)
```

Dep_time is sched_dep_time + dep_delay

**Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.**

```{r}
flights |>
  select(dep_time, dep_delay, arr_time, arr_delay)

flights |>
  select(starts_with("dep") & starts_with("arr"))

flights |> 
  select(dep_time:arr_delay, -contains("sched"))
```


**What happens if you specify the name of the same variable multiple times in a select() call?**

```{r}
flights |>
  select(dep_time, dep_time, dep_time)
```


IF you specify the name multiple times it does not affect and only prints the column once.

**What does the any_of() function do? Why might it be helpful in conjunction with this vector?**

```{r}
?any_of()
```

Any of does this "any_of() doesn't check for missing variables. It is especially useful with negative selections, when you would like to make sure a variable is removed.".

You can check if there are any of the arguments specified in any place.

```{r}
variables <- c("year", "month", "day", "dep_delay", "arr_delay")

flights |>
  select(any_of(variables))
```

**Does the result of running the following code surprise you? How do the select helpers deal with upper and lower case by default? How can you change that default?**

```{r}
flights |> select(contains("TIME"))

# to change this behaviour of ignoring the upper/lower

flights |>
  select(contains("TIME", ignore.case = FALSE))
```
It does suprise me because usually the upper case/lower case are not ignored, and it is a very important aspect. But in this case, the contains ignores it

**Rename air_time to air_time_min to indicate units of measurement and move it to the beginning of the data frame.**

```{r}
flights |>
  rename(air_time_min = air_time) |>
  select(air_time_min)
```

**Why doesn’t the following work, and what does the error mean?**

```{r}
flights |> 
  select(tailnum) |> 
  arrange(arr_delay)
#> Error in `arrange()`:
#> ℹ In argument: `..1 = arr_delay`.
#> Caused by error:
#> ! object 'arr_delay' not found
```

If you select just one column then you cannot arrange based on the variable that is not specified in that select()

## The pipe

THe pipe is like a nested function.

Previously, the pipe was %>% and not |>

## Groups

**group_by()**

Group into datasets
It is only visual when you do something else after the group_by

```{r}
flights |>
  group_by(month)
```

**summarize()**

This reduces the data frame to single row

```{r}
flights |>
  group_by(month) |>
  summarize(avg_delay = mean(dep_delay, na.rm = TRUE),
            n = n())

# n() returns the number of rows in each group
```

**slice_functions()**

slide_head(n = 1) # takes the first row from each group
slide_tail(n = 1) # takes the last row in each group
slide_min(x, n = 1) # takes the row with the smallest value of column x
slide_max(x, n = 1) #  takes the row with the largest value of column x
slide_sample(n = 1) # takes one random row

Instead of n you can use prop = 0.1 to select the 10% of the rows

Find the flights that are most delayed upon arrival at each destination
```{r}
# The with_ties = FALSE means that the program will just returns one row per group
flights |>
  group_by(dest) |>
  slice_max(arr_delay, n = 1, with_ties = FALSE) |>
  relocate(dest)
```

**Grouping by the multiple variables**

Creating groups with more than one variable

```{r}
daily <- flights |>
  group_by(year, month, day)
```

When you summarize a tibble grouped by more than one variable, each summary peels off the last group.

```{r}
daily_flights <- daily |> 
  summarize(n = n())

# to supress the message
# daily_flights <- daily |> 
#  summarize(
#    n = n(), 
#    .groups = "drop_last"
#  )
```

**Ungrouping**

```{r}
daily |>
  ungroup()
```

**.by**

Can be used to group the same way

```{r}

# Group a variable
flights |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE), 
    n = n(),
    .by = month
  )

# Or group by multiple variable
flights |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE), 
    n = n(),
    .by = c(origin, dest)
  )
```

### Exercise

**Which carrier has the worst average delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about flights |> group_by(carrier, dest) |> summarize(n()))**

```{r}
flights |>
  group_by(carrier, dest) |>
  summarize(delay = mean(dep_delay, na.rm = TRUE), n=n()) |>
  arrange(desc(delay))
```


**Find the flights that are most delayed upon departure from each destination.**

```{r}
# Shows the top 1 most departure delayed flights from each destination

flights |>
  group_by(dest) |>
  arrange(dest, desc(dep_delay)) |>
  slice_head(n = 1) |>
  relocate(dest, dep_delay)
```

**How do delays vary over the course of the day? Illustrate your answer with a plot.**

```{r}

delays_day <- flights |>
  group_by(hour) |>
  summarize(avg_delay = mean(dep_delay, na.rm = TRUE))

ggplot(delays_day, 
       aes(x = hour, y = avg_delay)) +
  geom_smooth(na.rm = TRUE)
```

**What happens if you supply a negative n to slice_min() and friends?**

```{r}
flights |>
  group_by(dest) |>
  slice_max(arr_delay, n = 1) |>
  relocate(arr_delay)
```

Using a negative value arranges the data frame in ascending or desensding order

**Explain what count() does in terms of the dplyr verbs you just learned. What does the sort argument to count() do?**

The count() counts the number of ibservations in each group, setting the sort argument to TRUE and arranges the categories in descending order.

**Suppose we have the following tiny data frame:**

```{r}
df <- tibble(
  x = 1:5,
  y = c("a", "b", "a", "a", "b"),
  z = c("K", "K", "L", "L", "K")
)
```

**Write down what you think the output will look like, then check if you were correct, and describe what group_by() does.**

```{r}
df |>
  group_by(y)
```

The output will be 5 different groups with 1, K. 2, K. 3, L...

**Write down what you think the output will look like, then check if you were correct, and describe what arrange() does. Also, comment on how it’s different from the group_by() in part (a).**

```{r}
df |>
  arrange(y)
```

This puts all a's first, with its corresponding values of x and z and then all the b's. It is different from the group_by() because there are no groups here

**Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does.**

```{r}
df |>
  group_by(y) |>
  summarize(mean_x = mean(x))
```

It separates in two groups of y (a and b) and it shows just the new variable called mean_x

```{r}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))
```

This creates the different groups that y and z can make: (a, K), (a, L) and (b, K) and the column called mean_x

```{r}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x), .groups = "drop")
```

It does the same thing just that it specifies that it just wants to show those variables that are used in the function

```{r}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))

df |>
  group_by(y, z) |>
  mutate(mean_x = mean(x))
```

The second one creates a new variable called mean_x. The difference between btoh is that the first only shows: y, z and mean_x and the second one shows also x

## Case study

Objective: l compare what proportion of times a player gets a hit (H) vs. the number of times they try to put the ball in play (AB)

```{r}
batters <- Lahman::Batting |> 
  group_by(playerID) |> 
  summarize(
    performance = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    n = sum(AB, na.rm = TRUE)
  )
batters
```

Now we plot the values

```{r}
batters |> 
  filter(n > 100) |> 
  ggplot(aes(x = n, y = performance)) +
  geom_point(alpha = 1 / 10) + 
  geom_smooth(se = FALSE)
```

With this we draw several conclusions: the variation in performance is larger with fewer opportunities to hit the ball and also that there's a positive correlation between performance and opportunities.

```{r}
batters |> 
  filter(n > 100) |> 
  ggplot(aes(x = n, y = performance)) +
  geom_point(alpha = 1 / 10) + 
  geom_smooth(se = FALSE)
```

