---
title: "Whole game - Data visualization"
output: html_notebook
---

# Whole game

## 1. Data visualization

With ggplot2

```{r}
# Loading the package

library(tidyverse)

# Data sets we will be using

library(palmerpenguins)
library(ggthemes)
```

### Penguins data frame

**Our ultimate job is to display the relationship between flipper lengths and body masses, taking into consideration the species of the penguin**

Variable -> Quantity, quality or property you can measure (variable refers to an attribute that all penguins have) 

 Value -> state of a variable

Oberservation ->  set of measurements made under similar conditions (all atributes of a single penguin) 

Tabular data -> set of values, each associated with a variable and an observation


```{r}
penguins
```

Variables as columns

```{r}
glimpse(penguins)
```
### Creating a ggplot

- mapping -> how variables are going to be mapped

- aes() -> define x and y axes

- geom -> geometrical object used to represent data

- color (inside of mapping) -> color based on the mapped points

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
       ) + 
# geompoint() is a scatterpoint
  geom_point(mapping = aes(color = species)) +
  geom_smooth(method = "lm")

```

*As the flipper length increases, so does the body mass. It is failry linear and strong*


