---
title: "Data transformation"
author: '@mortellss'
date: "2025-03-05"
output: html_document
---

Sometimes we want to: create new variables, rename ones that exist, reorder observations

```{r}
library(nycflights13)
library(tidyverse)
```


The nycflights13 package contains flights departed from NYC in 2013. The data comes from the Bureau of Transportation Statistics. This is a *tibble*. The difference between a tibble and a data frame is that tibbles are designed for large datasets so they only show the first few rows and only the columns that fit on one screen.

To see the data frames:
 
- View(flights) # Open a diffeerent tab with the data frame
- print(flights, width = Inf)  # To see all columns
- glimpse()



```{r}
flights
```

```{r}
?flights
```

### Rows

filter() -> changes which rows are present without changing their order
arrange() -> changes the order of rows without changing which are present
distinct() -> find rows with unique values

**filter()**

Changes the rows that are present without chaning their order

```{r}
# Flights that departed more than 120 minutes late
flights |>
  filter(dep_delay > 120)

# This is the same as:
# filter(flights, dep_delay > 120)
```

```{r}
# Flights that departed on January 1
flights |>
  filter(month == 1 & day == 1)

# Flights that departed on January or February

flights |>
  filter(month == 1 | month == 2)

# Another way is: 
# flights |>
#   filter(month %in% c(1, 2))


# You can also save it in a variable

jan1 <- flights |>
  filter(month == 1 & day == 1)


```

**arrange()**

Changes the order of rows based on the value of the columns.

```{r}
# This is in ascending order

flights |>
  arrange(year, month, day, dep_time)

# Descending order.
# ! Only allows one argument

flights |>
  arrange(desc(dep_time))
```
**distinct()**

Find all unique rows in dataet.

```{r}
flights |>
  distinct()

# Find all unique origin and destination pairs
# Just prints a table with that unique pairs
flights |> 
  distinct(origin, dest)


# .keep_all = TRUE
# 	Prints the table with those unique pairs but shows the entirety of the column

flights |>
  distinct(origin, dest, .keep_all = TRUE)
```

```{r}
# Find the number of ocurrences in that distinct
# The sort = TRUE, sorts it in descendent order

flights |>
  count(origin, dest, sort = TRUE)
```

### Exercises

**n a single pipeline for each condition, find all flights that meet the condition: **

```{r}
# Had an arrival delay of two or more hours

flights |>
  filter(arr_delay >= 120) |>
  arrange(desc(arr_delay))

# Flew to Houston (IAH or HOU).

flights |>
  filter(dest %in% c("IAH", "HOU"))

# Were operated by United, American, or Delta.

flights |>
  filter(carrier %in% c("UA", "AA", "DL"))

# Departed in summer (July, August, and September).

flights |>
  filter(month %in% c(7, 8, 9))

# Arrived more than two hours late but didn’t leave late.
# The view() opens up another tab with the dataset

flights |> 
  filter(arr_delay >= 120 & dep_delay <= 0) |> view()

# Were delayed by at least an hour, but made up over 30 minutes in flight

flights |>
  filter(dep_delay >= 60
         & (time_hour - air_time > 30))
```


**Sort flights to find the flights with the longest departure delays. Find the flights that left earliest in the morning.**

```{r}
flights |> 
  arrange(desc(dep_delay)) |>
  arrange(sched_dep_time) |>
  relocate(dep_delay, sched_dep_time)
  
```

The one with the longest departure delay

