---
title: "Workflow code style"
author: '@mortellss'
date: "2025-03-11"
output: html_document
---

# Workflow: code style

```{r}
install.packages("styler")
```
## Names

- Use lowercase letter, numbers and _
- It is better to use long, descriptive names
- The short ones are easy but are not descriptive.

## Spaces

- Put spaces on either side of mathematical operators apart from ^ (i.e. +, -, ==, <, …), and around the assignment operator
- Always put a space after a comma, just like in standard English.
- It's okay to add extra space

## Pipes
- Should always have a space before it and should typically be the last thing on a line
- If a function has several arguments, put each one of them on its own line

## Sectioning comments

Section to break up your file

### Exercise

**Restyle the following pipelines following the guidelines above.**

```{r}
flights |>
  filter(dest == "IAH") |> 
  group_by(year, month, day)|>
  summarize(n = n(),
  delay = mean(arr_delay, na.rm = TRUE)) |>
  filter(n > 10)
```

```{r}
flights |>
  filter(carrier == "UA",
  dest %in% c("IAH", "HOU"),
  sched_dep_time > 0900,
  sched_arr_time < 2000) |>
  group_by(flight) |>
  summarize(
    delay = mean(
      arr_delay, 
      na.rm=TRUE),
    cancelled = sum(is.na(arr_delay)),
    n=n()
  ) |>
  filter(n > 10)
```


