---
title: "Workflow basics"
author: '@mortellss'
date: "2025-03-05"
output: html_document
---

# Workflow: basics

## Coding basics

Creating variables.
 

```{r}
x <- 3 * 4
x
```
Assign a vector with multiple elements

```{r}
primes <- c(2, 3, 5, 7, 11, 13)
```

Arithmetics on vectors is applied to every element

```{r}
primes * 2
```

## Comments

In the code
```{r}
# This is a comment
```

Use comments to explain the why, not the what or how

## Names

- Separate words with _

```{r}
 this_is_a_very_long_name <- 2.5
```


- For autocompletion use: Ctrl + TAB


## Calling function

```{r}
seq(from = 1, to = 10)
```

```{r}
seq(1, 10)
```

- When R shows a + character it tells you that they are waiting for you to finish

### Exercises

**Why does this code not work?**

```{r}
my_variable <- 10
my_varıable
#> Error: object 'my_varıable' not found
```

It does not work because in the second line it uses a 1 instead or an i in the word variable

**Tweak each of the following R commands so that they run correctly:**


```{r}
library(tidyverse)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

**Press Option + Shift + K / Alt + Shift + K. What happens? How can you get to the same place using the menus?**

It shows you the shortcuts that R has